\textcolor{red}{Characterizing Water Quality with an Autonomous Robotic Team Employing Comprehensive In-Situ Sensing, Hyperspectral Imaging, Machine Learning, and Conformal Prediction}

### 2024-01-24 paper/paper.tex
Updates from Overleaf

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address[1]{%
Hanson Center for Space Sciences, University of Texas at Dallas, Richardson, TX 75080, USA;  John.Waczak@utdallas.edu (J.W.); Adam.Aker@utdallas.edu (A.A.); lhw150030@utdallas.edu (L.O.H.W.); Shawhin.Talebi@utdallas.edu  (S.T.); ashen.fernando@utdallas.edu  (B.F.); prabuddhamadusanka.hathurusinghedewage@utdallas.edu (P.H.); mazhar.iqbal@utdallas.edu (M.I.); mdlary@me.com~(M.D.L.); captdaveschaefer@gmail.com (D.S.); david.lary@utdallas.edu (D.J.L.)}
\corres{\hangafter=1 \hangindent=1.05em \hspace{-0.82em} Correspondence: {David.Lary@utdallas.edu}}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{\textcolor{red}{
Inland waters pose a unique challenge for water quality measurement by remote sensing techniques due to their complicated spectral features and small-scale variability. At the same time, collecting the high-quality reference data needed to calibrate remote sensing data products is both time-consuming and expensive. In this study, we present further development of a robotic team composed of an uncrewed surface vessel (USV) providing in-situ reference measurements and an unmanned aerial vehicle (UAV) equipped with a hyperspectral imager. Together, this team is able to address the limitations of existing approaches by enabling the simultaneous collection of comprehensive in-situ data together with high resolution hyperspectral imagery. We showcase the capabilities of this team using data collected in a North Texas pond during three observations in the Fall of 2020. Machine learning models for 13 parameters are trained using the dataset of paired in-situ measurements and coincident reflectance spectra. Using conformal prediction we  estimate 90\% confidence intervals for the output of each trained model. Maps generated by applying the models to the collected imagery reveal the small-scale spatial variability within the pond. Additionally, computing the permutation importance for each reflectance bin provides further insight into the relevant spectral features associated with each measurement.
}}

@@

Much effort has recently gone into the curation of comprehensive datasets combining water quality records with decades of satellite imagery to enable the development of new methods for water quality estimation. For example, Aurin et al. curated over 30 years of oceanographic field campaign data with associated coincident satellite imagery \cite{aurin2018remote}. Similarly, Ross et al. have combined over 600,000 records of dissolved organic carbon, chlorophyll-a, and other water quality parameters together with historical reflectance data from Landsat for the period of 1984-2019 \cite{ross2019aquasat}. With a sufficient quantity of data, machine learning methods can augment traditional approaches using reflectance band ratios to enable estimation of water quality metrics and perform water quality classification \cite{schiller1999neural, brown2008neural, lary2010artificial, feng2015urban,belgiu2016random}. However, the low spatial and spectral resolution of available multi-band remote sensing satellites makes it difficult to analyze inland waters with small features and complicated spectral signatures. Additionally, the long revisit times of these satellites limits their use for real-time monitoring, for example, during oil spills \cite{fingas2017review}.

@@

In this study, we expand on our previous work by demonstrating the capabilities of this coordinated robot team using data collected at a pond in North Texas across three separate days in the Fall of 2020. Specifically, we showcase the ability to accurately predict physical parameters such as temperature, conductivity, pH, and turbidity, as well as biochemical constituents including blue-green algae pigments, CDOM, and chlorophyll-a, in addition to concentrations of crude oils, optical brighteners, and a variety of ions. Further, the considerable size of the collected dataset allows us to quantify the uncertainty in the resulting machine learning models by using techniques from conformal prediction. This methodology addresses the shortcomings of existing approaches by facilitating rapid, real-time data collection coupled with on board hyperspectral image processing and machine learning model deployment with uncertainty quantification.

@@

In this paper, we have validated the methodology introduced in our previous work\cite{robotTeam1}. Specifically, we demonstrate that by using an autonomous team of sensing sentinels we can
\begin{enumerate}
    \item Rapidly acquire large volumes in in-situ reference data points for a variety of relevant water quality parameters including physical measurements such as temperature, pH, and conductivity as well as ion concentrations, chemicals such as crude oil, and relevant biochemical parameters such as blue-green algae pigments, chlorophyll-a, and CDOM. 
    \item Simultaneously collect hyperspectral remote sensing imagery using an UAV.
    \item Rapidly process the captured imagery into georectified reflectance data cubes.
    \item Train machine learning models to map captured reflectance spectra to the water quality parameters measured by the reference sensors.
    \item Apply the trained models to rapidly map large bodies of water and identify areas of interest.
\end{enumerate}
Additionally, we have demonstrated that we can extend this methodology in two new ways: First, we have shown that it is possible to combine the datasets from separate collection periods and successfully train models using \textit{all} of the records despite variability in incident lighting conditions that impact the assumptions made in the reflectance conversion of captured hyperspectral data cubes. Second, we have shown that by leveraging the large amount of collected data, we can simultaneously train the machine learning models and quantify their prediction uncertainties by utilizing conformal predictions.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{For research articles with several authors, a short paragraph specifying their individual contributions must be provided. The following statements should be used ``Conceptualization, X.X. and Y.Y.; methodology, X.X.; software, X.X.; validation, X.X., Y.Y. and Z.Z.; formal analysis, X.X.; investigation, X.X.; resources, X.X.; data curation, X.X.; writing---original draft preparation, X.X.; writing---review and editing, X.X.; visualization, X.X.; supervision, X.X.; project administration, X.X.; funding acquisition, Y.Y. All authors have read and agreed to the published version of the manuscript.'', please turn to the  \href{http://img.mdpi.org/data/contributor-role-instruction.pdf}{CRediT taxonomy} for the term explanation. Authorship must be limited to those who have contributed substantially to the work~reported.}

>>>>

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address[1]{%
Hanson Center for Space Sciences, University of Texas at Dallas, Richardson, TX 75080, USA;  John.Waczak@utdallas.edu (J.W.); Adam.Aker@utdallas.edu (A.A.); lhw150030@utdallas.edu (L.O.H.W.); shawhintalebi@gmail.com (S.T.); ashen.fernando@utdallas.edu  (B.F.); prabuddhamadusanka.hathurusinghedewage@utdallas.edu (P.H.); mazhar.iqbal@utdallas.edu (M.I.); mdlary@me.com~(M.D.L.); captdaveschaefer@gmail.com (D.S.); david.lary@utdallas.edu (D.J.L.)}
\corres{\hangafter=1 \hangindent=1.05em \hspace{-0.82em} Correspondence: {David.Lary@utdallas.edu}}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{\textcolor{red}{
Inland waters pose a unique challenge for water quality monitoring by remote sensing techniques due to their complicated spectral features and small-scale variability. At the same time, collecting the high-quality reference data needed to calibrate remote sensing data products is both time-consuming and expensive. In this study, we present the further development of a robotic team composed of an uncrewed surface vessel (USV) providing in-situ reference measurements and an unmanned aerial vehicle (UAV) equipped with a hyperspectral imager. Together, this team is able to address the limitations of existing approaches by enabling the simultaneous collection of comprehensive in-situ data together with high resolution hyperspectral imagery. We showcase the capabilities of this team using data collected in a North Texas pond across three observations in the Fall of 2020. Machine learning models for 13 parameters are trained using the dataset of paired in-situ measurements and coincident reflectance spectra. These models are able to successfully estimate physical parameters including temperature, conductivity, pH, and turbidity as well as the concentrations of blue-green algae, colored dissolved organic matter (CDOM), chlorophyll-a, crude oil, optical brighteners, and the ions $\mathrm{Ca}^{2+}$, $\mathrm{Cl}^{-}$, and $\mathrm{Na}^{+}$. We extend the training procedure to utilize conformal prediction in order to  estimate 90\% confidence intervals for the output of each trained model. Maps generated by applying each trained model to the collected imagery reveal the small-scale spatial variability within the pond. Additionally, the permutation importance computed for each feature of the trained models highlights the spectral features relevant to each water quality parameter.
}}

@@

Much effort has recently gone into the curation of comprehensive datasets combining water quality records with decades of satellite imagery to enable the development of new methods for water quality monitoring. For example, Aurin et al. curated over 30 years of oceanographic field campaign data with associated coincident satellite imagery \cite{aurin2018remote}. Similarly, Ross et al. have combined over 600,000 records of dissolved organic carbon, chlorophyll-a, and other water quality parameters together with historical reflectance data from Landsat for the period of 1984-2019 \cite{ross2019aquasat}. With a sufficient quantity of data, machine learning methods can augment traditional approaches using reflectance band ratios to enable estimation of water quality metrics and perform water quality classification \cite{schiller1999neural, brown2008neural, lary2010artificial, feng2015urban,belgiu2016random}. However, the low spatial and spectral resolution of available multi-band remote sensing satellites makes it difficult to analyze inland waters with small features and complicated spectral signatures. Additionally, the long revisit times of these satellites limits their use for real-time monitoring, for example, during oil spills \cite{fingas2017review}.

@@

In this study, we expand on our previous work by demonstrating the capabilities of this coordinated robot team using data collected at a pond in North Texas across three separate days in the Fall of 2020. Specifically, we showcase the ability to accurately predict physical parameters such as temperature, conductivity, pH, and turbidity, as well as biochemical constituents including blue-green algae pigments, CDOM, and chlorophyll-a, in addition to concentrations of crude oils, optical brighteners, and a variety of ions. Further, the considerable size of the collected dataset allows us to quantify the uncertainty in the resulting machine learning models by using techniques from conformal prediction. This methodology addresses the shortcomings of existing approaches by facilitating rapid, real-time data collection coupled with on board hyperspectral image processing and machine learning model deployment with uncertainty quantification. The resulting models are small enough to be deployed on board the UAV to enable real time streaming of data products to the ground. This approach drastically reduces the time from data collection to insight.

@@

In this paper, we have validated the methodology introduced in our previous work\cite{robotTeam1}. Specifically, we demonstrate that by using an autonomous team of sensing sentinels we can
\begin{enumerate}
    \item Rapidly acquire large volumes in in-situ reference data points for a variety of relevant water quality parameters including physical measurements such as temperature, pH, and conductivity as well as ion concentrations, chemicals such as crude oil, and relevant biochemical parameters such as blue-green algae pigments, chlorophyll-a, and CDOM. 
    \item Simultaneously collect hyperspectral remote sensing imagery using an UAV.
    \item Rapidly process the captured imagery into georectified reflectance data cubes.
    \item Train machine learning models to map captured reflectance spectra to the water quality parameters measured by the reference sensors.
    \item Apply the trained models to rapidly map large bodies of water and identify areas of interest.
\end{enumerate}
Additionally, we have demonstrated that we can extend this methodology in two new ways: First, we have shown that it is possible to combine the datasets from separate collection periods and successfully train models using \textit{all} of the records despite variability in incident lighting conditions that impact the assumptions made in the reflectance conversion of captured hyperspectral data cubes. Second, we have shown that by leveraging the large amount of collected data, we can simultaneously train the machine learning models and quantify their prediction uncertainties by utilizing conformal prediction.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{
Conceptualization, D.J.L.; methodology, J.L.W. and D.J.L; software, J.L.W; field deployment and preparation, J.L.W, A.A., L.O.H.W., S.T. B.F., P.H., M.I., M.L., D.S., and D.J.L.; validation, J.L.W.; formal analysis, J.L.W.; investigation J.L.W.; resources, D.J.L.; data curation, J.L.W., A.A., L.O.H.W., and D.J.L; writing---original draft preparation, J.L.W.; writing---review and editing, J.L.W. and D.J.L; visualization, J.L.W; supervision, D.J.L; project administration, D.J.L; funding acquisition, D.J.L.; All authors have read and agreed to the published version of the manuscript.
}
